Automatically Splitting a Staged Lambda Calculus

### 2015-07-09 paper/examples.tex
various edits

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare (k, n) of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except that the mixed-stage argument to both passes is
is explicitly masked, and we do not suppress the trivial first-stage result
returned by \texttt{qs1}.

>>>>

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare (k, n) of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except that the mixed-stage argument to both passes 
is explicitly masked, and we do not suppress the trivial first-stage result
returned by \texttt{qs1}.

### 2015-07-09 paper/semantics.tex
various edits

We express two-stage programs as terms in \lang, a typed, modal lambda calculus.

Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
and will be discussed in \ref{sec:needGround}.

@@

At the term level, {\tt next} blocks can be used to form future computations:
given a term $e$ of type $A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. 
This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
and it provides a handle (of type $\fut A$) now to that eventual value.
Computations at \bbonem\ can shuffle this handle around as a value, but the future result it refers to cannot be accessed.
This is because the only way to eliminate a $\fut$ wrapper is by using a {\tt prev}, which yields an $A$ at \bbtwo.
This feature was adapted from linear temporal logic, via \cite{davies96}.

$\curr A$ is a type in world \bbonem\ which classifies purely-first-stage
computations of type $A$. Given a world \bbonep\ term $e$ of type $A$,
$\pure{e}$ has type $\curr A$ at world \bbonem. ($e$ is guaranteed not to contain
second-stage computations because $\fut$ types are not available in world
\bbonep.)

An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
projection, of type $\curr A$.
This elimination form, in contrast to that of $\fut$, does not permit world \bbonem\ subterms
within any world \bbonep\ term.

Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
open on $\curr{A}$ and $\curr{B}$ respectively.
Without this, there would be no way to get information out from under a $\curr$ wrapper.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is monostage by definition, this second pass is just standard unstaged evaluation.
Consequently, this paper does not describe in any detail the second pass of evaluation.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
One consequence of the stronger requirement on partial values is that we will need a new kind of step to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val y = ...`
2`    val qs = y`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a function containing no first-stage redexes.

>>>>

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
and will be discussed in \ref{sec:needGround}.

@@

At the term level, {\tt next} blocks can be used to form future computations:
given a term $e$ of type $A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. 
This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
and it provides a handle (of type $\fut A$) now to that eventual value.
Computations at \bbonem\ can shuffle this handle around as a value, but the future result it refers to cannot be accessed.
This is because the only way to eliminate a $\fut$ wrapper is by using a {\tt prev}, which yields an $A$ at \bbtwo.
This feature was adapted from linear temporal logic, via \cite{davies96},
and ensures that there can be no flow of information from the second stage to the first.

$\curr A$ is a type in world \bbonem\ which classifies purely-first-stage
computations of type $A$. Given a world \bbonep\ term $e$ of type $A$,
$\pure{e}$ has type $\curr A$ at world \bbonem. ($e$ is guaranteed not to contain
second-stage computations because $\fut$ types are not available in world
\bbonep.)
An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
projection, of type $\curr A$.
This elimination form, in contrast to that of $\fut$, does not permit world \bbonem\ subterms
within any world \bbonep\ term.

Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we introduce the
$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
open on $\curr{A}$ and $\curr{B}$ respectively.
Importantly, this feature allows information to flow from \bbonep\ to \bbonem,
which would not be possible otherwise.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input top-level program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is monostage by definition, this second pass is standard unstaged evaluation
and is not described in further detail in this paper.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
\TODO Actually explain what MP and PE want this for.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
\TODO what?
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val y = ...`
2`    val qs = y`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a function containing no first-stage redexes.
